---
title: "Cyclistic Bike Share Analysis"
author: "Hrithik Sharma"
date: "2023-09-26"
output: html_document
---
## The Description

Three questions will guide the future marketing program:
1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members

## Prepring the Data
### Install and import Packages
```{r setup, include=FALSE}
# Set the CRAN mirror
options(repos = "https://cran.rstudio.com/")

#Install packages
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
install.packages("janitor")
install.packages("dplyr")

#import packages
library(tidyverse)
library(lubridate)
library(ggplot2)
library(janitor)
library(dplyr)
```

### Import all the Datasets
```{r}
#import all datasets
data_01 <- read.csv("202101-divvy-tripdata.csv")
data_02 <- read.csv("202102-divvy-tripdata.csv")
data_03 <- read.csv("202103-divvy-tripdata.csv")
data_04 <- read.csv("202104-divvy-tripdata.csv")
data_05 <- read.csv("202105-divvy-tripdata.csv")
data_06 <- read.csv("202106-divvy-tripdata.csv")
data_07 <- read.csv("202107-divvy-tripdata.csv")
data_08 <- read.csv("202108-divvy-tripdata.csv")
data_09 <- read.csv("202109-divvy-tripdata.csv")
data_10 <- read.csv("202110-divvy-tripdata.csv")
data_11 <- read.csv("202111-divvy-tripdata.csv")
data_12 <- read.csv("202112-divvy-tripdata.csv")
```

### Verify the datasets for same column names and data types
```{r}
#summarize the data frame
str(data_01)
colnames(data_01)
```

### Merge all the datasets in one after confirming that they have the same column names and data types
```{r}
#merge all data sets into one dataset
merged_data <- bind_rows(data_01,data_02,data_03,data_04,data_05,data_06,data_07,data_08,data_09,data_10,data_11,data_12)
```

### Verify the merged data
```{r}
head(merged_data)
colnames(merged_data)
str(merged_data)
```

## Process The Data
### Create new Data feilds for analysis
```{r}
#create date,month,day,year and day of week
merged_data$date <- as.Date(merged_data$started_at)
merged_data$month <- format(as.Date(merged_data$started_at), "%m")
merged_data$day <- format(as.Date(merged_data$started_at), "%d")
merged_data$year <- format(as.Date(merged_data$started_at), "%Y")
merged_data$day_of_week <- format(as.Date(merged_data$started_at), "%A")
```

### Inspect the new data fields
```{r}
# Inspect the new table that has been created
colnames(merged_data)  #List of column names
nrow(merged_data)  #How many rows are in data frame?
dim(merged_data)  #Dimensions of the data frame?
head(merged_data)  #See the first 6 rows of data frame.  Also tail(qs_raw)
str(merged_data)  #See list of columns and data types (numeric, character, etc)
summary(merged_data)  #Statistical summary of data. Mainly for numerics
```

### Convert the month from numeric to appropriate month names
```{r}
merged_data$month <- as.numeric(merged_data$month)
merged_data$month <- month.name[merged_data$month]
```

### Create a column to print the difference between the end time and start time
```{r}
#difference between end time and start time and have trip_duration in minutes
merged_data$ride_length <- difftime(merged_data$ended_at, merged_data$started_at)
is.factor(merged_data$ride_length)
merged_data$ride_length <- as.numeric(as.character(merged_data$ride_length))
is.numeric(merged_data$ride_length)
```

### Rename column names and modify data
```{r}
#Rename column name
colnames(merged_data)[colnames(merged_data) == "member_casual"] <- "usertype"
#rename member = subscriber and casual = customer
merged_data <- merged_data %>%
  mutate(usertype = ifelse(usertype == "member", "Subscriber", usertype)) %>% 
  mutate(usertype = ifelse(usertype == "casual", "Customer", usertype))

unique(merged_data$usertype)
```

### Delete 'BAD DATA' - delete data that has the Trip Duration in negative
```{r}
#delete data where trip_duration<0
clean_data_v1 <- merged_data[!(merged_data$ride_length<0),]
```

### Delete empty data
```{r}
#Delete empty data
clean_data_v1 <- na.omit(clean_data_v1)
head(clean_data_v1)
```

### Transform the trip_duration in mintues
```{r}
#ride_length converted in minutes for understandable analysis
clean_data_v1$ride_length_minutes <- clean_data_v1$ride_length / 60
```

## Analyise the Data
### Order the weekdays in the correct order
```{r}
clean_data_v1$day_of_week <- ordered(clean_data_v1$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

### Create an additional data frame
```{r}
add_data_3 <- clean_data_v1 %>% 
  mutate(weekday = wday(started_at, label = TRUE),
         month = month(started_at, label = TRUE),
         hour = hour(started_at)) %>%
  group_by(usertype, weekday, month, hour) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)
            ,rideable_type = first(rideable_type)) %>% 		# calculates the average duration
  arrange(usertype, weekday, month, hour)	
```

### Export the data
```{r}
write.csv(clean_data_v1, file = "D:/Data_Analysis/Cyclistic_2019_self_analysis/clean_data_v1.csv")
write.csv(add_data_3, file = "D:/Data_Analysis/Cyclistic_2019_self_analysis/add_data_v3.csv")
```

### Visualize data
Visualization is done in Tableau [Click Here](https://public.tableau.com/views/CyclisticBikeShareAnalysis_16956719298060/CyclisticBikeShareAnalysis?:language=en-US&:display_count=n&:origin=viz_share_link) to view the Visualization